---
title: "Cyclists bike sharing company"
author: "Saran B"
date: "2023-10-14"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

## Setting up my environment:
Notes: Setting up my environment by loading 'tidyverse' and 'lubridate' packages.

```{r loading packages}
library(tidyverse)
library(lubridate)
```

# Uploading the datasets:
Note: Here we are taking about 6 months different months data.
```{r loading dataset}
feb <- read.csv("C:\\Users\\saran\\OneDrive\\Documents\\Data_analysts_Case_study\\Cyclistic_company\\202302-divvy-tripdata.csv")
mar <- read.csv("C:\\Users\\saran\\OneDrive\\Documents\\Data_analysts_Case_study\\Cyclistic_company\\202303-divvy-tripdata.csv")
apr <- read.csv("C:\\Users\\saran\\OneDrive\\Documents\\Data_analysts_Case_study\\Cyclistic_company\\202304-divvy-tripdata.csv")
may <- read.csv("C:\\Users\\saran\\OneDrive\\Documents\\Data_analysts_Case_study\\Cyclistic_company\\202305-divvy-tripdata.csv")
jun <- read.csv("C:\\Users\\saran\\OneDrive\\Documents\\Data_analysts_Case_study\\Cyclistic_company\\202306-divvy-tripdata.csv")
jul <- read.csv("C:\\Users\\saran\\OneDrive\\Documents\\Data_analysts_Case_study\\Cyclistic_company\\202307-divvy-tripdata.csv")
```

# Checking whether all the column names in the different tables are same before combining into one:
```{r}
colnames(feb)
colnames(mar)
colnames(apr)
colnames(may)
colnames(jun)
colnames(jul)
```
# Checking the structure and datatype:
```{r}
str(feb)
str(mar)
str(apr)
str(may)
str(jun)
str(jul)
```

# Combining 6 month of data into a single data:
```{r}
ride <- rbind(feb,mar,apr,may,jun,jul)
```

# Inspecting the new table:
```{r}
colnames(ride) #checking the column names
nrow(ride) # checking how many rows are there
dim(ride) # checking the dimensions
str(ride) # Checking the columns datatypes
summary(ride) # Statistical summary of the data
head(ride) # checking first 6 data
tail(ride) # checking last 6 data
```

# Inserting NA to the blank cells:
```{r}
ride[ride == ""] <- NA
```

# Calculating the number of NA values in each columns:
```{r}
colSums(is.na(ride))
```

# Omiting the NA values in the dataset:
```{r}
ride_v2 <- na.omit(ride)
```

# Checking whether there is still null values in the dataset:
```{r}
colSums(is.na(ride_v2))
```

# Calculating the number of duplicated values:
```{r}
sum(duplicated(ride$ride_id))
```

# Calculating the number of unique values:
```{r}
length(unique(ride$ride_id))
```

# Converting datatype of  started_at and ended_at to date time format:
```{r}
ride_v2$started_at <- lubridate::as_datetime(ride_v2$started_at)
ride_v2$ended_at <- lubridate::as_datetime(ride_v2$ended_at)
```

# Checking whether th datatype is changed:
```{r}
class(ride_v2$started_at)
class(ride_v2$ended_at)
```

# Creating a new column for hour:
```{r}
ride_v2$start_hr <- lubridate :: hour(ride_v2$started_at)
ride_v2$end_hr <- lubridate :: hour(ride_v2$ended_at)
```

# Creating a new column for just date:
```{r}
ride_v2$start_date <- lubridate :: date(ride_v2$started_at)
ride_v2$end_date <- lubridate :: date(ride_v2$ended_at)
```

# Calculating the difference in time for ride length:
```{r}
ride_v2$ride_length <- difftime(ride_v2$ended_at,ride_v2$started_at)
ride_v2$ride_length <- as.numeric(ride_v2$ride_length)
```

# Calculating the difference in time for ride length in hours:
```{r}
ride_v2$ride_length_hr <- difftime(ride_v2$ended_at,ride_v2$started_at,units = c("hours"))
ride_v2$ride_length_hr <- as.numeric(ride_v2$ride_length_hr)
```

# Converting the dateyear format to weekday:
```{r}
ride_v2$start_day <- wday(ride_v2$start_date,label = TRUE)
ride_v2$end_day <- wday(ride_v2$start_date,label = TRUE)
```

# Checking the datatype and structure:
```{r}
str(ride_v2)
```

# Excluding the stationname 'Hq Qr' and ridelength less than or equal to 1:
```{r}
ride_v3 <- ride_v2[!(ride_v2$start_station_name == "HQ QR"|ride_v2$ride_length <=1),]
```

# Checking the dimension of the newly created dataset:
```{r}
dim(ride_v3)
```

# Conforming there is no null values in the newly created dataset:
```{r}
colSums(is.na(ride_v3))
```

# Taking a summary of the dataset:
```{r}
summary(ride_v3)
```

# Exporting the end result as CSV file to work on Visualization in Tableau:
```{r}
write.csv(ride_v3,"cyclist_bike_company.csv",row.names = FALSE)
```

 

